---
title: '#SEAInCodeRIntro'
author: "LaTreese Denson"
date: "11/23/2020"
output:
  word_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Workshop Goals:
Determine the influence of two environmental factors on species growth. Provide graphs and tables to summarize findings. Create a function to automate this analyses.

## Prerequisits;
You should have [R](https://cran.r-project.org) and [Rstudio](https://www.rstudio.com/products/rstudio/) installed. We will be working with R within the Rstudio environment.

# Determine, Set, and Create Directories
Determine which directory you are in:
```{r, echo = TRUE}
getwd()
```

**ProTip:** Use Ctrl + Enter (Windows) or Cmd + Enter (Mac) to run the current line of code using a keyboard shortcut.

Not the right directory? That's fine--let's change it!  
  
To change or set the directory in R use the `setwd()` function. 
```{r, eval=FALSE}
# The file path for your desktop
setwd("/Users/Me/Desktop")
```
**Note:** In this character string, you can only use a "/" or "\\\\" between folder names. 

For example, to get the directory to a specific folder:

1. From your desktop or start menu, navigate to the folder you wish to save in, and open it.
2. Right-click in any empty space inside the folder, select properties.
3. Select and copy the location (e.g., "C:\\\\Users\\\\Me\\\\Desktop").
4. Paste the location name inside the `setwd()` parentheses and add your folder name (e.g.,"\\\\SpecificFolder").

To keep things tidy, create another folder specifically for today's workshop. This can be done by creating the object `todays.file` where you can change the name whenever you wish.
```{r}
todays.file = "SEAInCode"
```

Use `dir.create()` to make the new folder then set the working directory to `todays.file`
```{r, eval = FALSE}
dir.create(todays.file)
setwd(todays.file)
```
**ProTip:** R does not like spaces. Use underscores instead.

Check to see if you are in the right place using `getwd()`, then use the function `dir()` to see what is in the folder.
```{r}
dir()
```
**ProTip:** Messed up? Want to move up one directory? No problem! Run `setwd("..")`.

# Install and Load Packages

There are millions of R packages designed to do all kinds of data analyses. Two packages that I have grown fond of for manipulating and visualizing data are `dplyr` and `ggplot2`. Fortunately, somebody decided to bundle these packages and a few others in what is called the "tidyverse". We can install this package using `install.packages()`. 

**dplyr:** a package for data manipulation

**ggplot2:** a package for creating graphics

**tidyverse:** the data analysis super package!

```{r, eval = FALSE}
# All necessary packages can be installed using this line
install.packages("tidyverse")
```
**Note:** Installing the package only needs to be done once.

Load the package every time a new R session is started using `library()`.
```{r, echo = TRUE}
library(tidyverse)
```
**Note:** If you are ever wondering how to use these packages check out the cheat sheets:

[ggplot2 cheat sheet](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf)

[dplyr cheat sheet](https://www.google.com/url?sa=t&rct=j&q=&esrc=s&source=web&cd=&ved=2ahUKEwjp7svH3pntAhVt0FkKHfR0DxYQFjAAegQIBBAC&url=http%3A%2F%2Fwww.rstudio.com%2Fwp-content%2Fuploads%2F2015%2F02%2Fdata-wrangling-cheatsheet.pdf&usg=AOvVaw2dHobUyR72zhFqfVO45D5A)

... or **Google it**!

# Data: Load, Explore,Visualize, Analyze, Vizualize again and Export Results

## Load Data
Read in the [growth_data](https://raw.githubusercontent.com/latreesedenson/SEAinCode/main/growth_data.csv) from Github using `read.csv()` and give it the object name "data":

```{r}
data = read.csv("https://raw.githubusercontent.com/latreesedenson/SEAinCode/main/growth_data.csv")

data=as.data.frame(data)
```
**Note:** You could also copy and paste the data from the web browser into an excel file, change the format to comma delimited, saved it as a csv file and then read it in using `read.csv()`. That way you would have a copy of the data on your local drive.

## Explore Data
Make sure the data was read in correctly using the function `head()`. 
```{r}
head(data)
```
**ProTip:** Not sure what a function does? Type `?head()` into the console and see what pops up in the help menu. Scroll down to the bottom--there are often examples of how to use the function that you can practice with.

What are the column names? Is everything there that should be there?
```{r}
names(data)
```

What are the dimensions of the dataset, how many rows and columns?
```{r}
dim(data)
```


Let's get some basic summary statistics from our data:minimum, maximum and mean.
```{r}
summary(data)
```


You will notice there are multiple species here and some NAs in the data. Let's work with the first species and remove the bad records.
We will use the pipe operator `%>%` from the magrittr package within the tidyverse package to do this. This syntax leads to code that is easier to write and to read. Use the keyboard shortcut: Ctrl + Shift + M (Windows) or Cmd + Shift + M (Mac) to get the pipe operator.

```{r}
Species_A_Clean = data %>% filter(species =="A",!is.na(light_hrs))

head(Species_A_Clean)
dim(Species_A_Clean)
summary(Species_A_Clean)
```


## Visualize
Let's do a boxplot of our response vs. our explanatory variables. Here we are using ggplot2 but I also provide code for using base R.
```{r}
ggplot(Species_A_Clean, aes(x=as.factor(light_hrs), y=growth)) + 
  geom_boxplot(fill="slateblue", alpha=0.2) +  xlab("light_hours")
ggsave("boxplot_light.tiff")

ggplot(Species_A_Clean, aes(x=as.factor(temp_range), y=growth)) + 
  geom_boxplot(fill="slateblue", alpha=0.2) +  xlab("temp")
ggsave("boxplot_temp.tiff")

# Base R
# tiff("boxplot.tiff")
# boxplot(growth~temp_range,data=Species_A_Clean,xlab="temp_range", ylab="growth(wt") 
# dev.off()
```

Attempt a histogram of the data using the `geom_histogram()` function. Change the `binwidth` to see what happens.
```{r}
ggplot(Species_A_Clean, aes(x = growth)) +
  geom_histogram(binwidth = 1)

ggsave("DataDistribution.tiff")

# Base R
# hist(Species_A_Clean$growth, breaks=1, col="darkgrey", main = "", xlab="growth") 
```


## Analyze
We use the `lm()` function which needs model and the data as it's input. The model is setup using a $\sim$, where $y\sim x$ means that Y is a function of X.

```r
lm(y ~ x1 + x2 + x3, Data = Data)

```
Let's run a model with our species data. We tell R that we have a categorical factor with multiple levels such as diet quality by using the function `factor()`.
```{r}
SpeciesA.lm = lm(growth ~ factor(light_hrs)*factor(temp_range), data = Species_A_Clean)
```

To get a summary of the linear model output we use the function `summary()`.
```{r}

sum.mod = summary(SpeciesA.lm)

####
# Additional functions you may find useful in the future
####
# coefficients(fit) # model coefficients
# confint(fit, level=0.95) # CIs for model parameters
# fitted(fit) # predicted values
# residuals(fit) # residuals
# anova(fit) # anova table 
####

results = sum.mod$coefficients

out = as.data.frame(results)

write.csv(out,"ModelCoefficeints.csv")

```
From our results we can conclude that for every XXX increase in density, this species specific growth decreases by 1 gram.

Let's plot this relationship and some diagnostics of the fit to the model.
```{r}
# tiff("ModelDiagnostics.tiff") # Create the graphic
# par(mfrow=c(2,2)) # set up how we want them to be plotted: 2 rows 2 columns
plot(SpeciesA.lm)
#dev.off()  # close the plotting window
```
This produces 4 main plots each used to diagnose the fit of the model to the data.

## Automation

Let' say you now need to do this for the two other species, nobody wants to write all that code out again. So now we will build our own function and if there is time we will loop through the species.

# Functions
Just like `plot()`, `header()`, `summary()`, `dim()`, etc. are all functions, we can make our own functions too.
Why do we make functions?

1.There is no available function to do what you want to do.

2.Work smarter -- not harder! You have a task that needs to be repeated, and you do not want to keep writing the same lines of code over and over again.

Functions must have a name and arguments, and they must return something. Arguments are the inputs you plan to use in your function. We often use the function `return()` to have our functions "print" to the console or pass something on to another function. Example layout:
```{r, eval = FALSE}
func.name = function(arg1,arg2) 
  { # bracket to start function
  
  # code to do something, here
  
  return (something)

  } # bracket to end function

```

Let's do the following in a function:
1. Select the species of interest
2. Produce plots exploring the response vs the explanatory variables
3. Produce plots of the data distribution
4. Fit the full linear model and export a table
5. Plot the diagnostics

```{r, eval=TRUE}
do.analysis = function(species.name = "B", input.data = data)
{
  species.data = input.data %>% filter(species == species.name)
  
  # put the boxplot stuff here
  ggplot(species.data, aes(x=as.factor(light_hrs), y=growth)) + 
    geom_boxplot(fill="slateblue", alpha=0.2) +  xlab("light_hours")
  ggsave("boxplot_light.tiff")
  
  ggplot(species.data, aes(x=as.factor(temp_range), y=growth))+
  geom_boxplot(fill="slateblue", alpha=0.2) +  xlab("temp")
  ggsave("boxplot_temp.tiff")
  
  final.mod = lm(growth~light_hrs+temp_range,data = species.data)
  
  sum.mod = summary(final.mod)
  
  results = sum.mod$coefficients
  
  out = as.data.frame(results)
  
  write.csv(out,"ModelCoefficeints.csv")
  
  tiff("modeldiagnostics.tiff")
  par(mfrow=c(2,2))
  plot(final.mod)
  dev.off()
  
  cat(paste("\nmodel outputs for species",species.name,"are complete!\n"))
}

# Try out your function
do.analysis("A",data)
```

# Loops
Create a vector of species names and perform the function on each species by looping the execution of our function
 
```{r,echo=FALSE}

species.list = c("A","B","C") # vector of species namesto loop through

for (i.species in 1:length(species.list))
{
  # i.species = 1 # test to see if your loop is working, it should change the current.species object
  current.species = species.list[i.species] # good practice to set an object name once and use it in multiple places
  dir.create(current.species) # create a new folder to work in or else things will overwrite because they have the same name
  setwd(current.species) # work in the newly created directory
  dir()
  do.analysis(current.species,data) # use the function we created previously
  setwd("..") # move back to the main folder to restart the loop
}
```

# Final ProTips
1. Comment your code! ( Use the hashtag # )
2. R is case-sensitive. For example, "head()" is not the same as "Head()." 
3. Be smart when naming files, folders, etc.. You will be looking for them later. Adding the date or sequential numbers to a folder name is helpful.
