---
title: 'PuttingTheSEAInCode'
author: "Chante Davis PhD & LaTreese Denson"
date: "12/04/2020"
output: 
  html_document:
    toc_float: true
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Workshop Goals:
- Determine the influence of two environmental factors on species growth. 
- Provide graphs and tables to summarize findings.
- Create a function to automate this analyses.

## Prerequisits:
You should have [R](https://cran.r-project.org) and [Rstudio](https://www.rstudio.com/products/rstudio/) installed. We will be working with R within the Rstudio environment.

# R script Setup
Determine which directory you are in:
```{r, echo = TRUE}
getwd()
```

**ProTip:** Use Ctrl + Enter (Windows) or Cmd + Enter (Mac) to run the current line of code using a keyboard shortcut.

Not the right directory? That's fine--let's change it!  
  
To change or set the directory in R use the `setwd()` function. 
```{r, eval=FALSE}
# Set the file path for your desktop
setwd("/Users/Me/Desktop")
```
**Note:** In this character string, you can only use a "/" or "\\\\" between folder names. 

For example, to get the directory to a specific folder:

1. From your desktop or start menu, navigate to the folder you wish to save in, and open it.
2. Right-click in any empty space inside the folder, select properties.
3. Select and copy the location (e.g., "C:\\\\Users\\\\Me\\\\Desktop").
4. Paste the location name inside the `setwd()` parentheses and add your folder name (e.g.,"\\\\SpecificFolder").

To keep things tidy, create another folder specifically for today's workshop. This can be done by creating the object `todays.file` where you can change the name whenever you wish. You can also make a string of file names or whatever you want using `c()`. 
```{r}
todays.file = "SEAInCode"
```
You can also make a string of file names or whatever you want using `c()`. This will come in handy later.

Use `dir.create()` to make the new folder then set the working directory to `todays.file`
```{r, eval = FALSE}
dir.create(todays.file)
setwd(todays.file)
```
**ProTip:** R does not like spaces. Use underscores instead.

Check to see if you are in the right place using `getwd()`, then use the function `dir()` to see what is in the folder.
```{r}
dir()
```
**ProTip:** Messed up? Want to move up one directory? No problem! Run `setwd("..")`.

# Install and Load Packages

There are millions of R packages designed to do all kinds of data analyses. Two packages that I have grown fond of for manipulating and visualizing data are `dplyr` and `ggplot2`. Fortunately, somebody decided to bundle these packages and a few others in what is called the "tidyverse". We can install this package using `install.packages()`. 

**dplyr:** a package for data manipulation

**ggplot2:** a package for creating graphics

**tidyverse:** the data analysis super package!

```{r, eval = FALSE}
install.packages("tidyverse")
```
**Note:** Installing the package only needs to be done once.

Load the package every time a new R session is started using `library()`.
```{r, warning=FALSE}
library(tidyverse)
```
**Note:** If you are ever wondering how to use these packages check out the cheat sheets:

[ggplot2 cheat sheet](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf)

[dplyr cheat sheet](https://www.google.com/url?sa=t&rct=j&q=&esrc=s&source=web&cd=&ved=2ahUKEwjp7svH3pntAhVt0FkKHfR0DxYQFjAAegQIBBAC&url=http%3A%2F%2Fwww.rstudio.com%2Fwp-content%2Fuploads%2F2015%2F02%2Fdata-wrangling-cheatsheet.pdf&usg=AOvVaw2dHobUyR72zhFqfVO45D5A)

... or **Google it**!

# Data Exploration

## Load Data
Read in the [growth_data](https://raw.githubusercontent.com/latreesedenson/SEAinCode/main/growth_data.csv) from Github using `read.csv()` and give it the object name "data" using `=` or `->` as the assignment operator:

```{r}
data = read.csv("https://raw.githubusercontent.com/latreesedenson/SEAinCode/main/growth_data.csv")

data = as.data.frame(data)
```
**Note:** You could also copy and paste the data from the web browser into an excel file, change the format to comma delimited, saved it as a csv file and then read it in using `read.csv()`. That way you would have a copy of the data on your local drive.

## Explore Data
Make sure the data was read in correctly using the function `head()`. 
```{r}
head(data)
```
**ProTip:** Not sure what a function does? Type `?head()` into the console and see what pops up in the help menu. Scroll down to the bottom--there are often examples of how to use the function that you can practice with.

What are the column names? Is everything there that should be there?
```{r}
names(data)
```

What are the dimensions of the dataset, how many rows and columns?
```{r}
dim(data)
```


Let's get some basic summary statistics from our data:minimum, maximum and mean.
```{r}
summary(data)
```


You will notice there are multiple species here and some NAs in the data. Let's work with the first species and remove the bad records.
We will use the pipe operator `%>%` from the magrittr package within the tidyverse package to do this. This syntax leads to code that is easier to write and to read. Use the keyboard shortcut: Ctrl + Shift + M (Windows) or Cmd + Shift + M (Mac) to get the pipe operator.

```{r}
Species_A_Clean = data %>% filter(species =="A",!is.na(temp_range))

head(Species_A_Clean)
dim(Species_A_Clean)
summary(Species_A_Clean)
```


## Visualize
Let's do a boxplot of our response vs. our explanatory variables. Here we are using the ggplot2 package, but I also provide code for using Base R.
You will notice that I have commented out some lines of code using '#'. This allows me to block out lines that I don't want to run, as well as create comments for my code.
```{r, echo=TRUE}
ggplot(Species_A_Clean, aes(x=as.factor(daylength), y=growth)) + 
  geom_boxplot(fill="slateblue", alpha=0.2) +  xlab("day length")

# To save to a file for viewing later we use the function ggsave()
# The .tiff extension is the type of file, you can also use .jpg
# .tiff files are good for publication graphics
# ggsave("boxplot_daylength.tiff")

ggplot(Species_A_Clean, aes(x=as.factor(temp_range), y=growth)) + 
  geom_boxplot(fill="slateblue", alpha=0.2) +  xlab("temp")

# ggsave("boxplot_temp.tiff")

# Here is the code to create a boxplot and save it in Base R
# Remove the # to run these 3 lines
# tiff("boxplot.tiff")
# boxplot(growth~temp_range,data=Species_A_Clean,xlab="temp_range", ylab="growth(wt")
# dev.off() 
```
Notice the way we save graphics is different depending on the package we are using. Also `dev.off()` closes all graphics devices so you can start a fresh graph next time.

Create a histogram of the data using the `geom_histogram()` function. Change the `binwidth` to see what happens.
```{r}
ggplot(Species_A_Clean, aes(x = growth)) +
  geom_histogram(binwidth = 1)

# You can remove the '#' in the line below if you want to save this
# ggsave("DataDistribution.tiff") 

# Base R
# hist(Species_A_Clean$growth, breaks=1, col="darkgrey", main = "", xlab="growth") 
```


## Analyze
We use the `lm()` function which needs model and the data as it's input. The model is setup using a $\sim$, where $y\sim x$ means that Y is a function of X.

```r
lm(y ~ x1 + x2 + x3, Data = Data)

```
Let's run a model with our species data. We tell R that we have a categorical factor with multiple levels such as diet quality by using the function `factor()`.
```{r}
SpeciesA.lm = lm(growth ~ factor(daylength)*factor(temp_range), data = Species_A_Clean)
```

To get a summary of the linear model output we use the function `summary()`.
```{r}

sum.mod = summary(SpeciesA.lm) 
# since sum.mod is an object 
# we will need to run the next line to see what is in it

sum.mod

####
# Additional functions you may find useful in the future
####
# coefficients(fit) # model coefficients
# confint(fit, level=0.95) # CIs for model parameters
# fitted(fit) # predicted values
# residuals(fit) # residuals
# anova(fit) # anova table 
####
```
Here you have your important outputs to make a conclusion with.

Let's save some of that information to an excel file. Note: you cannot save the entire output to a csv using the lines below. You may need some other functions.
```{r}
results = sum.mod$coefficients

out = as.data.frame(results)

write.csv(out,"ModelCoefficients.csv") # this is an excel file you can also write to a text file using write.table() and .txt file name instead of csv

```

Let's plot this relationship and some diagnostics of the fit to the model.
```{r}
# tiff("ModelDiagnostics.tiff") # Remember you can save this graphic
par(mfrow=c(2,2)) # set up how we want them to be plotted: 2 rows 2 columns
plot(SpeciesA.lm)
dev.off()  # close the plotting window and resets things
```
This produces 4 main plots each used to diagnose the fit of the model to the data.

# Automation
Let' say you now need to do this for the two other species, nobody wants to write all that code out again. Now we will build our own function and if there is time we will loop through and analyze the data for all of the species with the click of a button.

# Functions
Just like `plot()`, `header()`, `summary()`, `dim()`, etc. are all functions, we can make our own functions too.
Why do we make functions?

1.There is no available function to do what you want to do.

2.Work smarter -- not harder! You have a task that needs to be repeated, and you do not want to keep writing the same lines of code over and over again.

Functions must have a name and arguments, and they must return something. Arguments are the inputs you plan to use in your function. We often use the function `return()` to have our functions "print" to the console or pass something on to another function. Example layout:
```{r, eval = FALSE}
func.name = function(arg1,arg2) 
  { # bracket to start function
  
  # code to do something, here
  
  return (something)

  } # bracket to end function

```

Let's do the following in a function:
1. Select the species of interest
2. Produce plots exploring the response vs the explanatory variables
4. Fit the full linear model and export a table
5. Plot the diagnostics

```{r, eval=TRUE}
do.analysis = function(species.name = "B", input.data = data)
{
  # Use piping to select the data related to the species of choice in the first argument
  species.data = input.data %>% filter(species == species.name)
  
# Create boxplots and save them for publication
  ggplot(species.data, aes(x=as.factor(daylength), y=growth)) + 
    geom_boxplot(fill="slateblue", alpha=0.2) +  xlab("daylength")
  ggsave("boxplot_daylength.tiff")
  
  ggplot(species.data, aes(x=as.factor(temp_range), y=growth))+
  geom_boxplot(fill="slateblue", alpha=0.2) +  xlab("temp")
  ggsave("boxplot_temp.tiff")
  
  # Fit the model and save some of the results
  final.mod = lm(growth~daylength+temp_range,data = species.data)
  sum.mod = summary(final.mod)
  results = sum.mod$coefficients
  out = as.data.frame(results)
  write.csv(out,"ModelCoefficeints.csv")
  
  # Plot and save the diagnostics
  tiff("modeldiagnostics.tiff")
  par(mfrow=c(2,2))
  plot(final.mod)
  dev.off()
  
  # use cat() to print a message when you are done with your analyses
  cat(paste("\nmodel outputs for species",species.name,"are complete!\n"))
}

# Try out your function
do.analysis("A",data)
```

#  For Loops
There are various types of loops but "for" loops are most common. You use loops when you want to do something iteratively. For loops have the following structure:
```{r, eval = FALSE}
for (i in 0:5) # a variable and range of values it will take
  { # a bracket to start
   
  #some calculation or function to do something
  
  } # a bracket to end the for loop
```
Here "i" sequentially becomes a number from 0 to 5 and is used in each iteration of the loop.

Let's try a simple example where we want to add i to 5. Our output should consist of a list of values. We'll use the `cat()` function so we can see what the loop is doing. 
```{r}
for (i in 0:5) 
  {
  x = i + 5 # Do something to i, iteratively
  
  cat(x) # see what happens if you take this out
  
  # I like to print what I am doing inside of a loop ... 
  # uncomment the next line to see what it does
  
  # cat(paste("i= ", i, "and i + 5 =", x,"\n")) 
  
  } # end for loop
```

**Now for the ultimate test**
Use what we have learned so far to perform the analyses function on each species by using a loop.
 
```{r,warning=FALSE}

species.list = c("A","B","C") # vector of species names to loop through

for (i.species in 1:length(species.list))
{
  # i.species = 1 # test to see if your loop is working, it should change the current.species object
  current.species = species.list[i.species] # good practice to set an object name once and use it in multiple places
  dir.create(current.species) # create a new folder to work in or else things will overwrite because they have the same name
  setwd(current.species) # work in the newly created directory
  dir()
  do.analysis(current.species,data) # use the function we created previously
  setwd("..") # move back to the main folder to restart the loop
}
```

# Final ProTips
1. Comment your code! ( Use the hashtag # )
2. R is case-sensitive. For example, "head()" is not the same as "Head()." 
3. Be smart when naming files, folders, etc.. You will be looking for them later. Adding the date or sequential numbers to a folder name is helpful.
